The Development of a Collaborative Tool to Teach Debugging

### 2020-12-04 paper/paper.tex
Lots and lots of paper reorganization, section on the frontend started, intro modified

The goal of this project is to create a collaborative debugger to aid
in the teaching of debugging.  The collaborative debugger aims to help
teachers integrate this undertaught skill into their classes.  It
realizes the benefits of collaborative programming and teaching
debugging by providing a platform that is genuinely useful to students
and teachers.

@@

\subsection{Encouraging Collaboration}

@@

The collaborative debugger aims to provide a similar experience to
Replit.  Students should be able to start a debugging session, easily
invite other students, and interact with the debugger in a way that
makes it easy to share their knowledge.  By allowing multiple users to
interact with the same debugging session together in real-time, the
collaborative debugger lets students share experience and work through
problems together.  The debugger also makes it easy for teachers to
demonstrate debugging techniques remotely to multiple students, any of
whom can also interact with the debugger.  Since debugging sessions
are hosted remotely, users can start a session in one location and
resume it later from a different computer.
\par

@@

The aim is to provide students with an experience they can directly
reference when using rr or gdb in the future, as well as one which is
seamless to encourage the acquisition of debugging skills.

\subsection{Existing Tools that Enable Collaborative Programming} \label{exisitingcollab}

\section{Tools Used}

@@

  \includegraphics[scale=.8]{overall_system}
  \centering
  \caption{Overview of the Collaborative Debugger}
  \label{debugger:overview}
\end{figure}

The next sections give an overview of the various tools used to create
the collaborative debugger.  The debugger consists of:

\begin{enumerate}
\item A frontend web app built using React (\ref{react}) that presents
  a debugging interface to the end user.
\item A distributed backend managed by Kubernetes (\ref{k8s}), split
  into three parts:
  \begin{enumerate}
  \item A Pod for each debugging instance which runs the rr debugger
    (\ref{rr}).  These communicate directly with users through
    WebSocket server Pods. (\ref{socketio}).
  \item Pods running a frontend server written in Node.js which works
    in tandem with an API server created using Flask
    (\ref{flask/node}).  The API server manages creation and
    destruction of debug sessions, as well as authentication.
  \item A MongoDB (\ref{mongodb}) database.
  \end{enumerate}
\end{enumerate}

\subsection{Kubernetes}\label{k8s}

@@

\subsection{Mozilla's rr}\label{rr}

\subsubsection{Limitations}

\subsection{pygdbmi}

In order to ``support the development of systems which use the
debugger as just one small component of a larger system'', GDB
provides a machine-oriented interface called GDB/MI \cite{gdbman}. rr
supports interaction through GDB/MI, and using the interface was a
natural choice for the collaborative debugger.  In addition to being
far easier to interact with from within a program, the structured,
machine-friendly output of GDB/MI lends itself in particular to future
development of visualization aids in the collaborative debugger.
\par
To parse rr output into Python dictionaries and to easily control rr
as a subprocess, pygdbmi \cite{pygdbmi} is used in each debugging Pod.
pygdbmi's abstraction simplifies programatically controlling rr.  A
Pod can receive a command from the client, pass it to rr, and respond
without having to deal with parsing GDB/MI output or with directly
managing the rr process.

\subsection{Socket.IO}\label{socketio}

\subsection{MongoDB}\label{mongodb}

\subsection{Flask and Node.js}\label{flask/node}

\subsection{React}\label{react}

React is JavaScript library that simplifies creating user interfaces
and managing state \cite{react}.  React's state management is of
particular importance to the collaborative debugger's frontend.  State
constantly changes as users create/delete debugging sessions, join
existing sessions, and communicate with rr.  React allows classes to
encapsulate components such as a list of existing debugging sessions,
a view of the current program's source code, and the terminal
interface with rr.  Instances of these classes maintain state
and update efficiently.
\par
The frontend makes extensive use of JSX, syntax which allows the
inclusion of segments of HTML code within a React app written in
JavaScript.  This makes it easy for each component of the one-page
webapp to hide/show subcomponents as state changes.

\subsection{Monaco and Xterm.js}\label{xtermjs/monaco}

\section{Design}

The collaborative debugger consists of a distributed Kubernetes
backend and frontend React webapp.  Kubernetes was chosen for the
backend primarally so that a Pod could be created dynamically for each
debugging session.  The design of the backend is heavily distributed,
allowing individual components to be modified without the whole system
needing to be reconfigured.

@@

\begin{enumerate}
\item Statically created Pods.  These are the \textit{RR Message
    Server Pods}, the \textit{Frontend/API Server Pods}, and the
  \textit{Database Pod}.  This class of Pods are manually created when
  the cluster that will run the backend is first initialized.  The RR
  Message Server Pods and Frontend/API Server Pods may be created
  using Deployments \cite{k8s_docs} to allow later scaling, where
  multiple Pods running the same application may be created to
  facilitate increased load.
\item Dynamically created Pods.  This class of Pod contains the
  individual instances of \textit{RR Debug Session Pods} that are
  created on request by the API server.  When a client requests a new
  debug session, the API server uses the Kubernetes API to create a
  new Pod based on an existing template, gives the Pod a unique
  identifier, and associates it in the database with the requesting
  client.
\end{enumerate}

@@

Each component of the backend and frontend will be discussed in depth
in the following sections.

@@

\subsection{RR Debug Sessions \& RR Message Server}

@@

\subsubsection{RR Message Server}

@@

\subsubsection{RR Debug Session}

\paragraph{Building Example Pods} \label{buildingchannel}

@@

\begin{lstlisting}[language=Dockerfile,basicstyle=,caption={RR Debug Session Hash Example---Dockerfile},captionpos=b]
FROM example-container-registry.com/sproj/...

@@

\paragraph{The RR Translation Program} \label{joiningchannel}

@@

When the program starts, an instance of the \lstinline{RRInterface}
class, \lstinline{rri}, is initialized in the same scope as
\lstinline{sio}.  \lstinline{RRInterface} contains an instance of the
\lstinline{GDBController} class from pygdbmi to interface with rr and
control the rr subprocess, as well as a collection of methods to parse
rr output, the three most important of which are shown below:

@@

\subsection{Frontend/API Server}\label{api}

The second major backend system consists of the Frontend/API Server
and database.  This system is responsible for serving the frontend
webapp, as well as processing all API requests, such as those to
authenticate users and create new RR Debug Sessions.  The builtin
React development server is used to serve the frontend webapp and
redirect API requests to the Flask API server during development.  For
both security and performance reasons this should be changed to a
combination of a more robust solution like Nginx \cite{nginx} and a
production suitable web server for production.  The relationships
between components and overall process will remain unchanged after
this migration.
\par

@@

\subsubsection{Frontend/API Server Configuration}

@@

\subsubsection{The API Server}

@@

\paragraph{Processing a POST Request---Extracting Information}

@@

\paragraph{The Database} Before drilling into \lstinline{create_pod},
it may be helpful to examine the MongoDB database which
\lstinline{TranslationPodManager} and \lstinline{UserManager} interact
with.  The database consists of three collections: \lstinline{users},
\lstinline{pods}, and \lstinline{examples}.  Below is an example
record from each collection:

@@

\paragraph{Processing a POST Request---Dynamic Pod Creation}

@@

\paragraph{Processing a POST Request---Returning a Response}

@@

\subsection{Webapp}\label{webapp}

\section{Next Steps}

>>>>

Programmers spend large amounts of time debugging.  They often use
debuggers to assist them in this process of diagnosing sometimes
inexplicable problems with their code.  Though debuggers are primarily
used to hunt down hiden mistakes, they are also powerful tools to
understand program execution.
\par

The goal of this project is to create a collaborative debugger to aid
in the teaching of debugging.  The collaborative debugger hopes to
help teachers integrate this undertaught skill into their classes.  It
realizes the benefits of collaborative programming and teaching
debugging by providing a platform that aims to be useful to both
students and teachers.

@@

\subsection{Encouraging Collaboration} \label{exisitingcollab}

@@

The collaborative debugger aims to provide a similar experience to
tools like Replit.  Students should be able to start a debugging
session, easily invite other students, and interact with the debugger
in a way that makes it easy to share their knowledge.  By allowing
multiple users to interact with the same debugging session together in
real-time, the collaborative debugger lets students share experience
and work through problems together.  The debugger also makes it easy
for teachers to demonstrate debugging techniques remotely to multiple
students, any of whom can also interact with the debugger.  Since
debugging sessions are hosted remotely, users can start a session in
one location and resume it later from a different computer.
\par

@@

Currently, there are three example debug sessions included with the
collaborative debugger:

\begin{enumerate}
\item \lstinline{hash}: an example of classic pointer confusion.  The
  hashtable implementation included with the program is slightly
  bugged, where failing to call \lstinline{strncpy} results in the
  value for each key-value pair in the hash pointing to the same
  location in memory.  This is particuarly useful example for learning
  about setting watchpoints and breakpoints, as well as basic
  procedures to step through code.
\item \lstinline{smash}: a brief example of stack smashing.  A badly
  designed call to \lstinline{sscanf} results in a portion of the call
  stack being overwritten, and the process is killed as a result of
  stack smashing.  This example focuses on printing and examining
  memory locations.
\item \lstinline{memoization}: a program that demonstrates the
  difference between a recursive, iterative, and memoized recursive
  implimentation of the classic nth fibonacci number function.  This
  example focuses on examining the stack under conditions produced by
  different types of functions.
\end{enumerate}

\lstinline{memoization} will be covered in detail in the following
section on the frontend of the collaborative debugger (\ref{webapp}).
All three examples show how teachers can design debugging sessions
around specific concepts.  The collaborative debugger should make it
easy to distribute these sessions and for students to learn from them
collaboratively.
\par

Students using all three examples would benefit greatly from
visualization tools to help understand pointer locations, the process
space, and data structures.  The steps necessary to implement
visualization tools in the collaborative debugger are discussed
further in (\ref{next}).
\par

The aim of the collaborative debugger is to provide students with an
experience they can directly reference when using similar debuggers in
the future, as well as one which is seamless to encourage the
acquisition of debugging skills.  Creating such a platform will
hopefully encourage both students and teachers to explore the
undertaught area of debugging.

\section{Frontend Webapp}\label{webapp}

Before diving into the construction of the frontend webapp, it's
helpful to look at the steps two students would take to debug the
\lstinline{memoization} example program remotely.

@@

  \includegraphics[width=\textwidth]{memoization}
  \centering
  \caption{Debugging the Memoization Example Session}
  \label{debugger:overview}
\end{figure}

\begin{enumerate}
\item Students A and B both log into the collaborative debugger from
  different computers.
\item Student A creates a new Debug Session by clicking the button for
  \lstinline{memoization} under ``New Session''.  A few moments later,
  they're presented with the main screen of the debugger.
\item Using the session code (17734 in the example above), Student B
  joins the Debug Session.
\item Both students can now interact with the same debugger remotely.
  Input and output from the debugger and the current source file are
  synced between all students.  Students can work together to
  understand the reasons why memoization often improves the execution
  time of recursive programs.
\end{enumerate}

Notice that the debugging window in the example shown above is called
``rr''.  rr is an enhancment to gdb that is foundational in the
collaborative debugger's ability to deterministically record and
replay program execution. rr and its role in the collaborative
debugger will be covered extensively in (\ref{rrpods}).  Users
interact with rr through the collaborative debugger's frontend.

\subsection{Tools Used}

The frontend of the collaborative debugger must handle constant
changes in state as users participate in the process outlined above.
The frontend uses the following tools to manage state and display
output.

\subsubsection{React}\label{react}

React is JavaScript library that simplifies creating user interfaces
and managing state \cite{react}.  React's state management is of
particular importance to the collaborative debugger's frontend.  State
constantly changes as users create/delete debugging sessions, join
existing sessions, and communicate with rr.  React allows classes and
functions to encapsulate components such as a list of existing
debugging sessions, a view of the current program's source code, and
the terminal interface with rr.  Instances of these classes maintain
state and update efficiently.
\par

The frontend makes extensive use of JSX, syntax which allows the
inclusion of segments of HTML code within a React app written in
JavaScript.  This makes it easy for each component of the one-page
webapp to hide/show subcomponents as state changes.

\subsubsection{Monaco and Xterm.js}\label{xtermjs/monaco}

\section{Backend Design Overview}

The design of the collaborative debugger's backend is heavily
distributed, allowing individual components to be modified without the
whole system needing to be reconfigured.  Kubernetes in conjunction
with Docker are used to create the backend of the collaborative
debugger.

\subsection{Tools Used}

Kubernetes and Docker were chosen largely so that new Debug Sessions
could be easily and securely created on demand.

\subsubsection{Docker}\label{docker}

Docker is containerization software.  Docker containers encapsulate
applications and their dependencies while sharing their host's kernel.
This allows for lightweight, portable, and secure execution of complex
programs \cite{docker}.  All programs in the collaborative debugger's
backend run in Docker containers.  Docker containers' lightweight
nature and security are instrumental when trying to quickly create
containers to run user-defined code.

\subsubsection{Kubernetes}\label{k8s}

@@

\subsection{Overview}

The collaborative debugger consists of a distributed Kubernetes
backend and frontend React webapp.  Kubernetes was chosen for the
backend primarally so that a Pod could be created dynamically for each
debugging session.

@@

\begin{enumerate}
\item Statically created Pods.  These are the \textit{RR Message
    Server Pods}, the \textit{Frontend/API Server Pods}, and the
  \textit{Database Pod}.  This class of Pods are manually created when
  the cluster that will run the backend is first initialized.  The RR
  Message Server Pods and Frontend/API Server Pods may be created
  using Deployments \cite{k8s_docs} to allow scaling, where multiple
  Pods running the same application may be created to facilitate
  increased load.
\item Dynamically created Pods.  This class of Pod contains the
  individual instances of \textit{RR Debug Session Pods} that are
  created on request by the API server.  When a client requests a new
  debug session, the API server uses the Kubernetes API to create a
  new Pod based on an existing template, gives the Pod a unique
  identifier, and associates it in the database with the requesting
  client.
\end{enumerate}

@@

Each component of the backend will be discussed in depth in the
following sections.

@@

\section{RR Debug Sessions \& RR Message Server}

@@

\subsection{RR Message Server}

\subsubsection{Tools Used}

The collaborative debugger needs an effecient way for clients to
communicate with running RR Debug Sessions. The Socket.IO library was
selected to construct this efficient communication pathway.

\paragraph{Socket.IO}\label{socketio}

@@

\subsection{RR Debug Session} \label{rrpods}

\subsubsection{Tools Used}

\paragraph{Mozilla's rr}\label{rr}

\paragraph{pygdbmi}

In order to ``support the development of systems which use the
debugger as just one small component of a larger system'', gdb
provides a machine-oriented interface called gdb/mi \cite{gdbman}. rr
supports interaction through gdb/mi, and using the interface was a
natural choice for the collaborative debugger.  In addition to being
far easier to interact with from within a program, the structured,
machine-friendly output of gdb/mi lends itself in particular to future
development of visualization aids in the collaborative debugger.
\par
To parse rr output into Python dictionaries and to easily control rr
as a subprocess, pygdbmi \cite{pygdbmi} is used in each debugging Pod.
pygdbmi's abstraction simplifies programatically controlling rr.  A
Pod can receive a command from the client, pass it to rr, and respond
without having to deal with parsing gdb/mi output or with directly
managing the rr process.

\subsubsection{Building Example Pods} \label{buildingchannel}

@@

\begin{lstlisting}[language=Dockerfile,basicstyle=\linespread{0.5}\ttfamily,caption={RR Debug Session Hash Example---Dockerfile},captionpos=b]
FROM example-container-registry.com/sproj/...

@@

\subsubsection{The RR Translation Program} \label{joiningchannel}

@@

When the program starts, an instance of the \lstinline{RRInterface}
class, \lstinline{rri}, is initialized in the same scope as
\lstinline{sio}.  \lstinline{RRInterface} contains an instance of the
\lstinline{gdbController} class from pygdbmi to interface with rr and
control the rr subprocess, as well as a collection of methods to parse
rr output, the three most important of which are shown below:

@@

% \section{Tools Used}

% \begin{figure}[h!]

%   \includegraphics[scale=.8]{overall_system}
%   \centering
%   \caption{Overview of the Collaborative Debugger}
%   \label{debugger:overview}
% \end{figure}

% The next sections give an overview of the various tools used to create
% the collaborative debugger.  The debugger consists of:

% \begin{enumerate}
% \item A frontend web app built using React (\ref{react}) that presents
%   a debugging interface to the end user.
% \item A distributed backend managed by Kubernetes (\ref{k8s}), split
%   into three parts:
%   \begin{enumerate}
%   \item A Pod for each debugging instance which runs the rr debugger
%     (\ref{rr}).  These communicate directly with users through
%     WebSocket server Pods. (\ref{socketio}).
%   \item Pods running a frontend server written in Node.js which works
%     in tandem with an API server created using Flask
%     (\ref{flask/node}).  The API server manages creation and
%     destruction of debug sessions, as well as authentication.
%   \item A MongoDB (\ref{mongodb}) database.
%   \end{enumerate}
% \end{enumerate}

\section{Frontend/API Server}\label{api}

The second major backend system consists of the Frontend/API Server
and database.  This system is responsible for serving the frontend
webapp, as well as processing all API requests, such as those to
authenticate users and create new RR Debug Sessions.

\subsection{Tools Used}

\subsubsection{MongoDB}\label{mongodb}

\subsubsection{Flask and Node.js}\label{flask/node}

\subsection{Overview}

The builtin React development server is used to serve the frontend
webapp and redirect API requests to the Flask API server during
development.  For both security and performance reasons this should be
changed to a combination of a more robust solution like Nginx
\cite{nginx} and a production suitable web server for production.  The
relationships between components and overall process will remain
unchanged after this migration.
\par

@@

\subsection{Frontend/API Server Configuration}

@@

\subsection{The API Server}

@@

\subsubsection{Processing a POST Request---Extracting Information}

@@

\subsubsection{The Database} Before drilling into \lstinline{create_pod},
it may be helpful to examine the MongoDB database which
\lstinline{TranslationPodManager} and \lstinline{UserManager} interact
with.  The database consists of three collections: \lstinline{users},
\lstinline{pods}, and \lstinline{examples}.  Below is an example
record from each collection:

@@

\subsubsection{Processing a POST Request---Dynamic Pod Creation}

@@

\subsubsection{Processing a POST Request---Returning a Response}

@@

\section{Next Steps} \label{next}
